---
title: "Session 1"
output: html_document
date: "2024-01-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Session 1

## Importing Data, Data Wrangling, and Tidy Data

### Objectives of today's session

1.  Import various datasets

2.  Use `dplyr` package to preprocess our datasets

## Plan

1.  Read excel file, show arguments how they work
2.  Read tab-delimited file, show arguments how they work
3.  Read comma-sep files, show arguments how they work
4.  For other types of files, json, xml look for function in R
5.  **Exercise: Import a sample CSV file and an Excel file into R from directory**
6.  Introduction to **`dplyr`** packages.
7.  Common data manipulation tasks: filtering, selecting, mutating, summarizing.
8.  Grouping data and performing operations on groups.
9.  Joining data from different sources.
10. Handling missing data.
11. **Exercise: ask students to do some basic operations on sample dataset from prev exercise**

Tidy data is **a standard way of mapping the meaning of a dataset to its structure**.

```{r}
library(tidyr)
library(dplyr)
```

------------------------------------------------------------------------

## R is simple calculator

| Operator | Description                 |
|----------|-----------------------------|
| \+       | Addition                    |
| \-       | Subtraction                 |
| \*       | Multiplication              |
| /        | Division                    |
| \^       | Exponentiation (Power)      |
| %%       | Modulus (Remainder)         |
| %/%      | Integer Division (Quotient) |

```{r}
398 + 783

600 / 5

45 * 45
```

[**Exercise: Try doing following calculation using math operators**]{style="color:red"}

$$
\frac{1+5^{1/2}} {2}
$$

```{r}
# Your code here

```

[**Exercise: find modulus of 5 divided by 2**]{style="color:red"}

```{r}
# Your code here

```

------------------------------------------------------------------------

# R script

# Data types

```{r}
num_variable <- 42.5
class(num_variable)
```

```{r}
char_variable = "Welcome to the DS club"
class(char_variable)
```

```{r}
logic_variable = TRUE
class(logic_variable)
```

```{r}
num_variable == 42.5
class(num_variable == 42.5)
```

# R Functions

3 Key Properties of an R function.

-   functions have a name and they are case sensetive

-   following the name they have a pair of ()

-   inside (), function take 0 or more arguments

```{r}
summ_two_numbers <- function(a=3, b=6){
  return(a+b)
}

summ_two_numbers()
summ_two_numbers(5, 19)
```

[**Exercise: Complete this code to return square of 9**]{style="color:red"}

```{r}
# Your code here
# squared_num <- function(x){
#   squared = ___
#   return(___)
# }
# 
# squared_num(9)
```

------------------------------------------------------------------------

# R vectors

```{r}
nums_vector = c(2, 5, 10, 60)

nums_vector[3]

nums_vector[2:3]

nums_vector[nums_vector > 9]
```

[**Exercise: create a vector of odd numbers till 10 and return numbers greater than 5**]{style="color:red"}

```{r}
# Your code here

```

# Read data

```{r}
# UNCOMMENT these lines
# df = read.table('file.txt')
# df = read.csv('file.csv')
```

# Using libraries

All libraries are installed in this environment

```{r}
# UNCOMMENT these lines
# library(readr)
# 
# df = readr::read_csv('file.csv', col_select = c('col1'))
```

[Exercise: install library for excel files "readxl" and read "file.xlsx" using `read_excel` function]{style="color:red"}

```{r}
# UNCOMMENT these lines
# library(readxl)
# excel_file = readxl::read_excel('file.xls', sheet = 'Sheet1')
# excel_file
```

------------------------------------------------------------------------

# Split cells with `tidyr` package

[Exercise: load `tidyr` package]{style="color:red"}

```{r}
# Let's create sample dataframe and look at some tidyr functions to merge and split data

splcing_data <- data.frame(
  Event_ID = c('ENSG171863_RI', 
               'ENSG168356_SE', 
               'ENSG160185_SE', 
               'ENSG243960_A5',
               'ENSG197291_A3'),
  PSI_value_1 = as.character(c('0.', '0.', '0.', '0.', '1.')),
  PSI_value_2 = as.character(c(1,2,3,4,0)),
  Group = c("A", "B", "B", "A", "A"), 
  Attribute1 = c('NA', 5, "NA", 9, 'NA')
)

splcing_data
```

```{r}
new_splicing_data = tidyr::unite(splcing_data, PSI_value_1, PSI_value_2, col="PSI_value", sep="", remove=TRUE)
new_splicing_data

# return new dataframe with new defined column
# also we can remove old columns or remain them
# you could play with "sep" and "col" arguments
```

```{r}
# tidyr::separate same as tidyr::separate_wider_delim
tidyr::separate(splcing_data, Event_ID, sep="_", into=c('ID', 'Splicing event'), remove=TRUE)
```

[Exercise: work with another dataset, unite and separate this dataset]{style="color:red"}

```{r}
# Your code here

```

[Exercise: Unite all columns of ***splicing_data*** into one column and named it ***all_columns***]{style="color:red"}

```{r}
# Your code here

```

# Reshape data

```{r}
# First of all, let's create sample dataframe and then reshape it
table2 <- data.frame(
  country = c('A', 'A', 'A', 'A', 'B', 'B', 'B', 'B', 'C', 'C', 'C', 'C'),
  year = c(1999, 1999, 2000, 2000, 1999, 1999, 2000, 2000, 1999, 1999, 2000, 2000),
  type = c('cases', 'pop', 'cases', 'pop', 'cases', 'pop', 'cases', 'pop', 'cases', 'pop', 'cases', 'pop'),
  count = c('0.7K', '19M', '2K', '20M', '37K', '172M', '80K', '174M', '212K', '1T', '213K', '1T')
)

table2
```

```{r}
# Let's reshape it
tidyr::pivot_wider(table2, names_from = type, values_from = count)
```

[**I think we don't need exercise here**]{style="color:red"}

# Dplyr package

# joining, transforming and summarizing

### Return back to our splicing data, and join it to another dataframe with new Attribute2, but new dataframe has information only for 4 Events

What to do??? ---\> left and right joins

```{r}
new_splicing_data$PSI_value = as.numeric(new_splicing_data$PSI_value)
new_splicing_data
```

```{r}
tmp_df = data.frame(
  Event_ID = c('ENSG171863_RI', 'ENSG168356_SE', 'ENSG160185_SE', 'ENSG243960_A5'),
  Attribute2 = c(43, 54, 72, 26))
tmp_df
```

### dplyr::left_join

```{r}
dplyr::left_join(new_splicing_data, tmp_df, by="Event_ID")
```

### dplyr::right_join

```{r}
dplyr::right_join(new_splicing_data, tmp_df, by="Event_ID")
```

### dplyr::select

```{r}
# Select needed columns 
new_splicing_data %>% select(Event_ID)
```

### dplyr::mutate()

```{r}
# Transform data in dataframe
new_splicing_data %>% mutate(new_PSI_value = 10*PSI_value)
```

### dplyr::filter()

```{r}
new_splicing_data %>% filter(PSI_value > 0.2)
```

### dplyr::group_by() & dplyr::summarise()

```{r}
new_splicing_data %>% 
  group_by(Group) %>%
  summarise(
    mean_value = mean(PSI_value)
  )
```

[**Exercise: for below splicing data**]{style="color:red"}

[1. **Select Events with PSI_value below 0.5**]{style="color:red"}

[2. **For found Events, find sum and mean values for each groups where these Events are**]{style="color:red"}

```{r}
new_splicing_data

# Your code here
# ...
```

# Write data

```{r}
write.csv(new_splicing_data, 'new_splicing_data.csv')
 
# readr::write_excel_csv()
```
